Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-28 overleaf/paper_SIMD.tex
Edits to 1 and 2.

\section{Introduction}
\label{sec:introduction}

Multi-party computation (MPC) allows $N$ parties $p_1,\dots, p_N$ to perform a computation on their private inputs securely. Informally, security means that the secure computation protocol computes the correct output (correctness) and it does not leak any information about the individual party inputs, other than what can be deduced from the output (privacy).

% The first results from the early 80's~\cite{FOCS:Yao82b,STOC:GolMicWig87,STOC:BenGolWig88,STOC:ChaCreDam88} introduced the problem and formally defined its security, and investigate feasibility in different models---e.g., passive or malicious adversaries, unconditional or computational security, with or without a broadcast channel. This spawned a rich theoretical literature of beautiful works improving and extending in various manners the original ideas.

MPC theory dates back to the early 1980-ies~\cite{FOCS:Yao82b,STOC:GolMicWig87,STOC:BenGolWig88,STOC:ChaCreDam88}. 
Long the realm of theoretical cryptography, MPC has seen significant advances in programming technology in recent years. 
These advances bring MPC closer to practice and wider applicability --- 
MPC technology has been employed in real-world scenarios such as auctions~\cite{FC:BCDGJK09}, biometric identification~\cite{ESORICS:BlaGas11}, 
and privacy-preserving machine learning~\cite{SP:MohZha17,CCS:MohRin18}. 
The goal is to bring the technology to a level where programmers can write \emph{secure} and \emph{efficient} programs without 
commanding extensive knowledge of cryptographic primitives.

The problem, therefore, is to build a high-level programming language and a compiler, and there has been significant advance in 
this space, e.g., ~\cite{CCS:BenNisPin08,ESORICS:BogLauWil08,CCS:ZhaSteBla13,Hastings:2019,Keller:2020,Acay:2021,Braun:2022} among other work. 
Current research largely falls at the two ends 
of the classical compiler: (1) work on \emph{front-end} language design and (2) work on \emph{back-end} protocol implementation. Work on language design 
focuses on high-level constructs necessary to express multiple parties, computation by different parties, and information flow from one party 
to another~\cite{SP:RasHamHic14, Acay:2021}. On the other end, work on protocol implementation focuses on cryptographic foundations and their efficient circuit-level
implementation~\cite{NDSS:DemSchZoh15, CCS:ABFKLO18, Braun:2022}, e.g., implementation of operations (e.g., MUL, ADD) using different sharing protocols 
(Boolean or Arithmetic GMW~\cite{STOC:GolMicWig87} or Yao's garbled circuits~\cite{FOCS:Yao82b}), as well as efficient share conversion from one representation 
to another.

Earlier compilers did both back-end and front-end translation without a specific focus on either, as their aim was to demonstrate applicability of MPC on real-world
programming problems. As the field advanced, works have focused more closely on front-end language design (e.g., Wysteria~\cite{SP:RasHamHic14} and Viaduct~\cite{Acay:2021}) 
or back-end ``circuit-level'' design and implementation (e.g., MOTION~\cite{Braun:2022}).

In this work we focus on an intermediate language and what we call \emph{backend-independent optimizations}, in a close analogy to \emph{machine-independent} optimizations
in the classical compiler. The following figure summarizes our key idea:

{\begin{center}
\includegraphics[width=0.8\linewidth]{figs_paper_SIMD/focus.png}
\end{center}
}

We emphasize the MPC Source intermediate representation and optimization over MPC Source. As in classical compilers, we envision different front ends (e.g., Wysteria,
our front end IMP-MPC) compiling into MPC Source. MPC Source is particularly suitable for optimizations such as protocol mixing~\cite{Ishaq:2019, Fang:2022} and SIMD-vectorization, 
which takes advantage of amortization at the circuit level. The MPC Source IR exposes the linear structure of MPC programs, which simplifies program analysis; this is
in contrast to source, which has if-then-else constructs. In the same time, MPC Source is sufficiently ``high-level''  to support analyses and optimizations that take into account 
control and data flow in a specific program. MPC Source is small in size and analysis is tractable, as opposed to analysis over an unrolled circuit~\cite{Ishaq:2019}. Again as
in classical compilers, we envision translation of MPC Source (optimized or unoptimized) into MOTION, SPDZ, or other back-end code.

%In contrast, the high-level program contains if-then-else and array writes, or the low-level circuit . 
%The compilation problem, i.e., translation from a high-level language to circuits, is difficult and therefore, breaking the problem into sub-problems can  tractable.

\subsection{Our Contribution} In this paper, we develop a compiler framework that takes a Python-like routine and produces MOTION code: we describe (a) the IMP-MPC language, its syntax and semantic restrictions, (b) translation from IMP-MPC into MPC Source, (c) a specific backend-independent optimization: novel SIMD-vectorization on MPC Source, and (d) translation from MPC Source into MOTION code. %\ana{I think we have to add more on MOTION, specifically mixing and amortization.}

We focus on the MOTION framework as our back-end for several reasons. First, it is the state of the art in terms of performance~\cite{Braun:2022}. Second, it provides an API over efficient implementation for a wide variety of cryptographic operations in three different protocols ---  Arithmetic GMW, Boolean GMW, and BMR --- which allows for protocol mixing~\cite{Ishaq:2019, Fang:2022}, a known backend-independent optimization. Third, MOTION provides API for SIMD-level operations, which amortize cost and lead to significant improvement in memory footprint and 
throughput~\cite{NDSS:DemSchZoh15, CCS:ABFKLO18, Braun:2022}. It also enables MPC Source-level optimizations such SIMD vectorization.

Our second contribution is an analytical model for cost estimation of amortized schedules. Originally, we hoped that optimal scheduling (under our model) was tractable, as the problem
appeared simpler than the classical scheduling problem. Unfortunately, we show that optimal scheduling is NP-hard via a reduction to the Shortest Common Supersequence (SCS) problem. 
Cost estimation is important as it drives not only vectorization but other optimizations such as protocol mixing optimizations and scheduling as well~\cite{Ishaq:2019, Fang:2022}.

Our most important contribution is the implementation and evaluation of the compiler framework. We demonstrate expressivity of the source language by running the compiler on \ana{X} programs; these include classical MPC benchmarks such as PSI and Biometric matching, as well as kMeans, Histogram, and others. Vectorization leads to \ana{X}x improvement in number of gates on average, \ana{Y}x speedup in circuit generation time, \ana{Z}x speedup in the LAN setting, and \ana{Z}x speedup in the WAN setting using the Boolean GMW protocol. Results form BMR are similar.

Our results emphasize the importance of backend-independent optimizations --- vectorization (described in this work) and protocol mixing (tackled in previous works~\cite{CCS:BDKKS18,Ishaq:2019, Fang:2022}) are two optimizations readily available at the level of MPC Source. We believe that our work can lead to future work on backend-independent compilation and optimization, bringing in new MPC optimizations and combinations of optimizations, much in the vein of standard compilers, and bring MPC programming technology closer to practice and wider applicability.

\subsection{Outline}

The rest of the paper is organized as follows. \secref{sec:overview} presents an overview of the compiler. \secref{sec:model} describes our model for cost estimation and argues NP-hardness of optimal scheduling and \secref{sec:compiler} describes the compiler. \secref{sec:experiments} presents implementation and evaluation. \secref{sec:related} discusses related work and \secref{sec:conclusion} concludes.

\begin{comment}
\begin{itemize}
\item We define the scheduling problem for MPC. 
We present an analytical model to reason about cost of schedules
and show that scheduling is NP-hard via a reduction to the shortest 
common supersequence problem.

\item We present a compiler that takes an IMP-like high-level 
program and produces amortized (i.e., vectorized) low-level cryptographic 
code in the MOTION framework. Central contributions are 1) a novel compiler framework, 
2) a vectorization algorithm that produces optimal schedules for a large 
number of MPC programs, and 3) reasoning over output arrays without Array SSA; we remove 
infeasible loop-carried dependences introduced by standard SSA to improve vectorization.

\item We present an implementation and evaluation in the MOTION framework. 
\ana{Fill in with final results. Mention benchmarks (standard + new ones + HyCC). }

\end{itemize}
\end{comment}

\section{Overview}
\label{sec:overview}

\begin{table*}
\begin{tabular}{ccc}
\begin{minipage}{0.275\textwidth}
{\small
\begin{pythonn}
def biometric(C: shared[list[int]], D: int, 
      S: shared[list[int]], N: int) -> 
      shared[tuple[int,int]]:
   min_sum : int = MAX_INT
   min_idx : int = 0  
   for i in range(N): 
      sum : int = 0
      for j in range(D): 
         # d = S[i,j] - C[j]
         d : int = S[i * D + j] - C[j] 
         p : int = d * d
         sum = sum + p
      if sum < min_sum:
         min_sum : int = sum
         min_idx : int = i
   return (min_sum, min_idx)
\end{pythonn}
}
\end{minipage}

&

\begin{minipage}{0.33\textwidth}
{\tiny
\begin{pythonn}
min_sum!1 = MAX_INT
min_idx!1 = 0
for i in range(0, N):
   min_sum!2 = PHI(min_sum!1, min_sum!4)
   min_idx!2 = PHI(min_idx!1, min_idx!4)
   sum!2 = 0
   for j in range(0, D):
      sum!3 = PHI(sum!2, sum!4)
      d = SUB(S[((i * D) + j)],C[j]) 
      p = MUL(d,d) 
      sum!4 = ADD(sum!3,p) 
   t = CMP(sum!3,min_sum!2) 
   min_sum!3 = sum!3
   min_idx!3 = i
   min_sum!4 = MUX(t, min_sum!3, min_sum!2) 
   min_idx!4 = MUX(t, min_idx!3, min_idx!2) 
return (min_sum!2, min_idx!2)   
\end{pythonn}
}
\end{minipage}

&

\begin{minipage}{0.33\textwidth}
{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_idx!1 = 0
# S^ is same as S. C^ replicates C N times: 
S^ = raise_dim(S, ((i * D) + j), (i:N,j:D)) #S^[i,j] = S[i,j]
C^ = raise_dim(C, j, (i:N,j:D)) #C^[i,j] = C[j]

sum!2[I] = [0,..,0]
# computes _all_  "at once"
d[I,J] = SUB_SIMD(S^[I,J],C^[I,J])
p[I,J] = MUL_SIMD(d[I,J],d[I,J])

for j in range(0, D):
   # sum!2[I], sum!3[I], sum!4[I] are size-N vectors 
   # computes N intermediate sums "at once"
   sum!3[I] = PHI(sum!2[I], sum!4[I])       
   sum!4[I] = ADD_SIMD(sum!3[I],p[I,j])

min_idx!3[I] = [0,1,...N-1]   
for i in range(0, N):
   min_sum!2 = PHI(min_sum!1, min_sum!4) 
   t[i] = CMP(sum!3[i],min_sum!2)
   min_sum!4 = MUX(t[i], sum!3[i], min_sum!2)  
for i in range(0, N):
   min_idx!2 = PHI(min_idx!1, min_idx!4)    
   min_idx!4 = MUX(t[i], min_idx!3[i], min_idx!2)
return (min_sum!2, min_idx!2)   
\end{pythonn}
}
\end{minipage}

\\

(a) IMP Source & (b) MPC Source & (c) Optimized MPC Source
\end{tabular}
\caption{Biometric Matching: From IMP Source to MPC Source to Optimized MPC Source. }
\label{tab:source_and_MPC_source_and_optimized_MPC_source}
\end{table*}

\subsection{Source}

As a running example, consider Biometric matching, a standard MPC benchmark.
An intuitive (and naive) implementation is as shown in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(a).
Array \python{C} is the feature vector of \texttt{D} features that we wish to match and \python{S} 
is the database of \python{N} size-\python{D} vectors that we match against.

\begin{comment}
{\small
\begin{verbatim}
def biometric(C: shared[list[int]], D: int, 
              S: shared[list[int]], N: int) -> 
                      tuple[shared[int], shared[int]]:
    min_sum = 10000
    min_index = 0  
    for i in range(N): #loop over database
        sum = 0
        for j in range(D): #loop over features
            d = S[i * D + j] - C[j] #i.e., d = S[i,j] - C[j]
            p = d * d
            sum = sum + p
        if sum < min_sum:
            min_sum = sum
            min_index = i
    return (min_sum, min_index)
\end{verbatim}
}
\end{comment}

Our compiler takes essentially standard IMP~\ana{Need citation here.} 
syntax and imposes certain semantic restrictions. (We detail the restrictions in the following sections.)
The programmer writes an iterative program and annotate certain inputs 
and outputs as \emph{shared}. In the example arrays \python{C} and \python{S}
are shared, meaning that they store shares, however, the array sizes \python{D} and
\python{N} respectively are plaintext. The code iterates over the entries in the 
database and computes the Euclidean distance (its square actually) of the current 
entry \python{S[i]} and \python{C}. The program returns the index of the vector that gives 
the best match plus the corresponding sum of squares.

\subsection{MPC Source and Cost of Schedule}

Our compiler generates an intermediate representation, MPC Source. MPC Source 
for Biometric Matching is shown in 
Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(b). First, MPC Source is an SSA form. 
Second, it is linear. The conditional in lines 13-14 in IMP Source turns into the linear code in lines 12-16 in MPC Source. 
The test turns into a CMP operation (\python{t = CMP(sum!3,min\_sum!2)}), followed by the 
true-branch sequence, followed by the MUX operations. The first MUX operation selects the value
of \python{min\_sum}: if \python{t} is true, then \python{min\_sum} gets the value of the second multiplexer
argument,  \python{min\_sum!3}, otherwise it takes the value of the third argument, \python{min\_sum!2}.  
Third, MPC Source is a special form of SSA. The SSA $\phi$-nodes at the if-then-else (lines 13-15) turn into 
MUX operations, while the $\phi$-nodes at for loops turn into what we call pseudo PHI nodes with a straightforward semantics.

\begin{comment}
{\small
\begin{pythonn}
1. min_sum!1 = 10000
2. min_index!1 = 0
3. for i in range(0, N!0):
4.    min_sum!2 = PHI(min_sum!1, min_sum!4)
5.    min_index!2 = PHI(min_index!1, min_index!4)
6.    sum!2 = 0
7.    for j in range(0, D!0):
8.      sum!3 = PHI(sum!2, sum!4)
9.      d!3 = (S!0[((i * D!0) + j)] - C!0[j]) // MPC
10.     p!3 = (d!3 * d!3) // MPC
11.     sum!4 = (sum!3 + p!3) // MPC
12.  !1!2 = (sum!3 < min_sum!2) // MPC
13.  min_sum!3 = sum!3
14.  min_index!3 = i
15.  min_sum!4 = MUX(!1!2, min_sum!3, min_sum!2) // MPC
16.  min_index!4 = MUX(!1!2, min_index!3, min_index!2) // MPC
17.!2!1 = (min_sum!2, min_index!2)   
\end{pythonn}
}
\end{comment}

%The compiler linearizes the source turning conditionals into MUX statements. The PHI nodes
%are remnants of the SSA IR; the compiler generates code that picks the correct value 
%when producing MOTION output; the MOTION framework in turn linearizes 
%loops when it generates the circuit.

We turn to our analytical model to compute the \emph{cost} of the program. Assuming fixed 
cost $\beta$ for a local MPC operation (essentially just ADD) and cost $\alpha$ for a remote
MPC operation (e.g., MUX, CMP, and remaining operations), the cost of the iterative schedule 
will be $ND(2\alpha+\beta) + N(3\alpha$).

A key contribution is the vectorizing transformation. We can compute all $N*D$ 
subtraction operations (line 9) in a single SIMD instruction; similarly we can compute 
all multiplication operations (line 10) in a single SIMD instruction. And while we cannot
vectorize computation of the N individual sums, we can compute the N sums in parallel. 
Our compiler \emph{automatically detects these opportunities and transforms the program}. 
It is standard that MPC researchers write vectorized versions of the Biometric
program by hand; we are the first (to the best of our knowledge) to automatically 
transform a naive, iterative MPC program into an unintuitive vectorized one.

\subsection{Vectorized MPC Source and Cost of Schedule}

Our compiler produces the vectorized program shown in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(c). 
Note that this is still our intermediate representation, Optimized MPC Source. Subsequently,
the compiler turns this code into MOTION variables, loops and SIMD primitives, which MOTION then 
uses to generate the circuit.

The compiler determines that SUB and MUL in ``naive'' MPC Source (lines 9 and 10 in (b))
can be fully vectorized into the SIMD SUB and MUL in optimized MPC Source (lines 9 and 10 in (c)). 
Notation \python{p[I,J]} denotes a 2-dimensional array with fully vectorized dimensions.
The computation of sum (line 11 in (b))
is sequential across the $j$-dimension, but it is parallel across the $i$-dimension. 
The loop in lines 12-16 in (c) illustrates; here \python{p[I,j]} refers to the $j$-th column in \python{p}. 
Unfortunately, CMP and MUX remain sequential.

\begin{comment}
{\small
\begin{verbatim}
min_sum!1 = 10000
min_index!1 = 0
// S!0^ is same as S!0. C!0^ replicates C!0 N-times: 
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

sum!2 = [0,..,0]
// Computes all differences and all products "at once"
d!3[I,J] = SUB_SIMD(S!0^[I,J],C!0^[I,J])
p!3[I,J] = MUL_SIMD(d!3[I,J],d!3[I,J])

for j in range(0, D!0):
  // sum!2[I], sum!3[I], sum!4[I] are vectors of size N  
  // Computes N intermediate sums "at once"
  sum!3[I] = PHI(sum!2[I], sum!4[I])       
  sum!4[I] = ADD_SIMD(sum!3[I],p!3[I,j])

min_index!3 = [0,1,...N!0-1]

for i in range(0, N!0):
  min_sum!2 = PHI(min_sum!1, min_sum!4) 
  !1!2[i] = CMP(sum!3[i],min_sum!2)
  min_sum!4 = MUX(!1!2[i], sum!3[i], min_sum!2)

for i in range(0, N!0):
  min_index!2 = PHI(min_index!1, min_index!4)  
  min_index!4 = MUX(!1!2[i], min_index!3[i], min_index!2)
!2!1 = (min_sum!2, min_index!2)   
\end{verbatim}
}
\end{comment}

In MPC backends, executing $n$ operations ``at once'' in a single SIMD operation costs a lot less than executing those $n$ operations one by one.
This is particularly important when there is communication (i.e., in remote), since many 1-bit values are sent at once rather than sequentially. 
%In MPC compilers a vectorized operation that computing M operations "at once" costs 
%essentially the same ($\alpha$ or $\beta$) as an individual operation. 
We elaborate on the cost model in the following section but for now consider that 
each operation has a \emph{fixed} portion (does benefits from amortization) and 
a \emph{variable} portion (does not benefit from amortization): $\alpha = \alpha_\mathit{fix} + \alpha_\mathit{var}$. 
This gives rise to the following formula for amortized cost: $f(n) = \alpha_\mathit{fix} + n\alpha_\mathit{var}$, 
as opposed to unamortized cost $g(n) = n\alpha_\mathit{fix} + n\alpha_\mathit{var}$. (We extend the same reasoning to 
$\beta$-instructions.)

Thus, the fixed cost of the vectorized program amounts to
$2\alpha_\mathit{fix}$ + $D \beta_\mathit{fix} + N(3\alpha_\mathit{fix})$. (The variable cost is the same in both the vectorized and non-vectorized programs.) 
The first term in the sum corresponds to the vectorized
subtraction and multiplication, the second term corresponds to the FOR loop on $j$ and the third 
one corresponds to the remaining FOR loops on $i$. Clearly, $2\alpha_\mathit{fix} + D\beta_\mathit{fix} + N(3\alpha_\mathit{fix}) << ND(2\alpha_\mathit{fix}+\beta_\mathit{fix}) + N3\alpha_\mathit{fix}$.
Empirically, we observe 10x to 50x improvement over un-vectorized Biometric Matching in circuit generation time and setup time and 5x to 30x improvement in online time.
Additionally, the un-vectorized version runs out of memory for $N=256$, while the vectorized one runs with the standard maximal input size $N=4,096$.
\ana{Edit numbers with final experiments.}

\section{Analytical Model}
\label{sec:model}

\ana{We need an intro to section here.}

\subsection{Scheduling in MPC}
\label{sec:mpc}

%\ana{Here we need more on MPC before jumping to scheduling. MPC-source, basic assumptions about static loop bounds, etc.}

For this treatment we make the following simplifying assumptions:

\begin{enumerate}
\item All statements in the program execute using the same protocol (sharing). That is, there is no share conversion. 
%\ishaq{This is not an assumption, this is our setting. We work in the single protocol world.}
\item There are two tiers of MPC instructions, local and remote. A local instruction (essentially just ADD) has cost $\beta$ 
and a remote instruction (e.g., MUX, MUL, SHL, etc.) has cost $\alpha$, where $\alpha >> \beta$. We assume that all remote 
instructions have the same cost.
%\ishaq{Don't we need to distinguish between local (cheap) vs. interactive (expensive) instructions? I don't see how we could assume this. -- Past suggestion from Vassilis: Ideally we should assume cost to be a convex function.}\ana{After discussion 1/12: replace single costs 1 with two levels of costs: $\alpha$ for non-local operations, e.g., MUL, and $\beta$ for local ops, e.g., ADD.}
\item  In MPC frameworks, executing $n$ operations ``at once'' in a single SIMD operation costs a lot less than executing those $n$ operations one by one.
Following Amdahl's law~\ana{citation!}, we write $\alpha = \frac{1}{s}p\alpha + (1-p)\alpha$, where $p$ is the fraction of execution time that benefits from amortization and $(1-p)$ 
is the fraction that does not, and $s$ is the available resource. Thus, $n\alpha = \frac{n}{s}p\alpha + n(1-p)\alpha$. 
For the purpose of the model we assume that $s$ is large enough and the term $\frac{n}{s}p\alpha$ amounts to a \emph{fixed cost} incurred regardless of
whether $n$ is $10,000$ or just $1$. (This models the cost of sending a packet from party A to party B for example.) The bottom line is that amortized execution 
of $n$ operations is $f(n) = \alpha_\mathit{fix} + n\alpha_{var}$ in contrast to unamortized execution $g(n) = n\alpha_\mathit{fix} + n\alpha_{var}$ and, of course,
$\alpha_\mathit{fix} << n\alpha_\mathit{fix}$.
%We assume infinite parallel capacity---i.e., a single MPC-instruction costs as much as $N$ amortized instructions, namely $\alpha$ or $\beta$. 
%This is a standard assumption in Cryptographic Parallel RAM. ABY presents empirical support for this assumption~\ana{Add citations. PRAM, ABY.}
%\ishaq{Comment from Vassilis: We assume there are infinite parallel capacities, this is the standard assumption in PRAM (Cryptographic Parallel RAM).}\ana{PRAM assumption stays. We have to replace the unit cost of 1 with a suitable amortization function $f(n)$, which I don't think changes much in the cost modeling and analysis.}
\item MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
2 MUL instructions can be amortized in a single SIMD instruction that costs $\alpha_\mathit{fix} + 2\alpha_\mathit{var}$, however a MUL and a MUX instruction 
still cost $2\alpha_\mathit{fix} + 2\alpha_\mathit{var}$ even when scheduled ``in parallel''. 
%MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
%2 MUL instructions scheduled in parallel benefit from amortization and cost $\alpha$, however a MUL and a MUX instructions scheduled
%in parallel still cost $2\alpha$. 
%\ishaq{We need to reword this assumption to something like "$K$ parallel MUL costs much less than $K$ (because they get amortized), but any mix of $K$  MUL and MUX still cost roughly $K$. Specifically, we should take away the low constants (2 and 1) because we know for low constants this is not true.} \ana{Again, core assumption that MUL and MUX don't benefit from amortization stays. We have to replace constant costs with functions, as in (3).}
\end{enumerate}

\input{prob_stmt_np_hardness.tex}

\begin{figure*}
  \includegraphics[width=1.0\linewidth]{figs_paper_SIMD/compiler_framework.png}
  \caption{Compiler Framework. \ana{Change figure since we are not doing div-and-conquer.}}
  \label{fig:compiler_framework}
\end{figure*}

\section{\bf Compiler Framework}
\label{sec:preliminaries}

%We assume arbitrarily nested loops in the MPC-source IR and read-only arrays. We assume that loops range from 0 to some constant N. Arrays are linearized (row-major order as in MOTION) and accesses are via functions of the induction variables of the enclosing loops.

\figref{fig:compiler_framework} presents an overview of our compiler. This section describes the front-end phases of the compiler
and Section~\secref{sec:vectorization} describes our vectorization algorithm. \ana{We'll need a section on MOTION code gen too.}
%\ana{Need to state this more precisely.}

\subsection{Syntax and Semantic Restrictions}
\label{sec:syntax}

Source syntax is essentially standard IMP syntax as shown below:
%\begin{figure}[t]
{\small
\[
\begin{array}{l@{~~~~~}l}
  \begin{array}{l@{~}l@{~~~}l}
 e & ::= e \; \mathit{op} \; e \mid e \mid \x \mid  \texttt{const} \mid \texttt{A[$e$]} & \mathit{expression} \\
 % s &::= t \;\f & \mathit{field} \\
 %\md &::= t\;{\m}(t \;{\this}, \; t\;\x) \; \{\; \ol{t\;\y}\;\s;\;\return\;\y \; \} 
 %                 & \mathit{method} \\ 
  s & ::= s;s \mid  & \mathit{sequence} \\
  & \x = e \mid \texttt{A[$e$]} = e \mid & \mathit{assignment \; stmt} \\
  & \texttt{for $i$ in range($I$)}: s \mid & \mathit{for \; stmt} \\
  & \texttt{if $e$: $s$ else: $s$} & \mathit{if \; stmt} \\ 
%   t &::= q \; {\sf C} & \mathit{qualified \; type}\\
%   q &::= \high \mid \poly \mid \low & \mathit{FlowCFL \; qualifier}\\
%    t' &::= q' \; \C & \mathit{field \; qualified \; type}\\
%   q' &::= \tainted \mid \poly & \mathit{field \; qualifier}\\

  \end{array}
\end{array}
\]
}
%\end{figure}

The syntax allows for array accesses, arbitrarily nested loops, and if-then-else control flow. 
Expressions are typed $q \; \tau$, where qualifier $q$ and base type $\tau$ are:

{\small
\[
\begin{array}{l@{~~~~~}l}
  \begin{array}{l@{~}l@{~~~}l}
 \tau & ::= \texttt{int}  \mid \texttt{bool} \mid \texttt{list[int]} \mid  \texttt{list[bool]} & \mathit{base \; types} \\
  q & ::= \texttt{shared} \mid \texttt{plain} & \mathit{qualifiers} \\
  \end{array}
\end{array}
\]
}

The type system is mostly standard, and in our experience, a sweet spot between readability and expressivity.
The \texttt{shared} qualifier denotes shared values, i.e., ones shared among the parties and computed upon 
under secure computation protocols; the \texttt{plain} qualifier denotes plaintext values. Shared lists denote 
shared elements, but plaintext length. Subtyping is again as expected: $q_1 \; \tau <: q_2 \; \tau$ 
iff $q_1 <: q_2$ where $\texttt{plain} <: \texttt{shared}$. Thus, we can convert a plaintext value into a shared one, 
but not vice versa.

Our compiler imposes certain semantic restrictions that it enforces throughout the various 
phases of compilation. We note that in some cases, the restrictions 
can be easily lifted and we plan to do so in future iterations of the work.

\begin{enumerate}
\item Loops are of the form $0 <= i < I$ and bounds are fixed at compile time. 
It is a standard restriction in MPC that the bounds must be known at circuit-generation time.
\item Arrays are one-dimensional. N-dimensional arrays are linearized and accessed 
in row-major order and at this point the programmer is responsible for linearization
and access. (This restriction can be easily lifted.)
\item Array subscrpts are plaintext values as specified by the rule:
\begin{semantics}
    \ntyperule{Array Access}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e : \texttt{plain int} \quad \Gamma \vdash \texttt{A} : q \; \texttt{list}[\tau] \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{A[$e$]} : q \; \tau
}
\end{semantics}
The subscript $e$ is a function of the indices of the enclosing loops. 
For read access, the compiler allows an arbitrary such function. 
However, it restricts write access to \emph{canonical writes}, i.e., \texttt{A[i,j,k]} = ...
where \texttt{i}, \texttt{j} and \texttt{k} loop over the three dimensions of \texttt{A}.  
Write access such as for example \texttt{A[i,j+2]} = ... is not allowed. 
%\item \ana{Add rules for logical ops}
%\item \ana{Add rules for arithmetic ops}
\item The final restriction involves MUX as expressed by the rule: 
\begin{semantics}
    \ntyperule{MUX}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e_1 : q_1 \; \texttt{bool} \quad \Gamma \vdash e_2 : q_2 \; \tau \quad \Gamma \vdash e_2 : q_2 \; \tau \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{MUX($e_1$,$e_2$,$e_3$)} : q_1 \vee q_2 \vee q_3 \; \tau
}
\end{semantics}
\end{enumerate}
Importantly, the arguments of MUX are restricted to base types, as passing array arguments 
is not valid MOTION MPC. \ana{Check this.} This causes a slight inconvenience as we could not write 
\begin{pythonn}
if e: A[i] = val
\end{pythonn}
Instead we had to write
\begin{pythonn}
if not(e): val = A[i]
A[i] = val
\end{pythonn}

We note that while the programmer writes 
annotations at the level of IMP Source (recall \ana{ref to overview section}), the 
annotations propagate through the transformations and are checked at the 
level of MPC Source.

%\ana{Disadvantage of MUX}

For the rest of this section we write $i,j,k$ to denote the loop nest: $i$ is the outermost loop, $j$, is immediately nested in $i$, and so on until $k$
and we use $I,J,K$ to denote the corresponding upper bounds. We write $A[i,j,k]$ to denote canonical access 
to an array element. In the program, canonical access is achieved via the standard row-major order formula: $(J*K)*i + K*j + k$.
To simplify the presentation we describe our algorithms in terms of three-element tuples $i,j,k$, however, discussion easily generalizes to 
arbitrarily large loop nests.

\subsection{From IMP Source to SSA}

Our compiler translates from Source to SSA as follows:

\begin{enumerate}
  \item
        \textbf{Parsing}:
        Use Python's \texttt{ast} module to parse the input source code to a Python AST.
  \item
        \textbf{Syntax checking}:
        Ensure that the AST matches the restricted subset defined in Section~\secref{sec:syntax}.
        This step outputs an instance of the \texttt{restricted\_ast.Function} class, which represents our restricted subset of the Python AST.
  \item
        \textbf{3-address CFG conversion}:
        %\benjamin{TODO: Is this a good amount of detail?} \ana{This is good amount of detail for the Tech report. In the actual paper we'll shorten a lot and add pointers to the Tech report.}

        Convert the restricted-syntax AST to a three-address control-flow graph.
        To do this, first,
        add an empty basic block to the CFG and mark it as current.
        Next, for each statement in the restricted AST's function body,
        process the statement.
        Statements can either be for-loops, if-statements, or assignments (as in~\secref{sec:syntax}).
        Rules for processing each kind of statement are given below:
        \begin{enumerate}
          \item
                \textbf{For-loops}:
                Create new basic blocks for
                the loop condition
                (the \emph{condition-block}),
                the loop body
                (the \emph{body-block}),
                and the code after the loop
                (the \emph{after-block}).
                Insert a jump from the end of the current block to the condition-block.
                Then, mark the condition-block as the current block.
                Insert a for-instruction at the end of the current block with the loop counter variable and bounds from the AST.
                Next, add an edge from the current block to the after-block labeled ``FALSE'' and
                an edge from the current block to the
                body-block labeled ``TRUE''.
                Then, set the body-block to be the current block
                and process all statements in the AST's loop body.
                Finally, insert a jump to the condition-block and set the after-block as current.
          \item
                \textbf{If-statements}:
                Create new basic blocks for
                the ``then'' statements of the if-statement
                (the \emph{then-block}),
                the ``else'' statements of the if-statement
                (the \emph{else-block}),
                and the code after the if-statement
                (the \emph{after-block}).
                At the end of the current block,
                insert a conditional jump to the then-block or else-block
                depending on the if-statement condition in the AST.
                Next,
                mark the then-block as current,
                process all then-statements,
                and add a jump to the after-block.
                Similarly,
                mark the else-block as current,
                process all else-statements,
                and add a jump to the after-block.
                Finally, set the after-block to be the current block,
                and give it a \emph{merge condition} property equal to the condition of the if-statement.
          \item
                \textbf{Assignments}:
                In the restricted-syntax AST,
                the left-hand side of assignments
                can be a variable or an array subscript.
                If it is an array subscript, e.g., \texttt{A[i] = x},
                change the statement to \texttt{A = Update(A, i, x)}.
                If the statement is not already three-address code,
                for each sub-expression in the right-hand side of the assignment,
                insert an assignment to a temporary variable.
        \end{enumerate}
  \item
        \textbf{SSA conversion}:
        Convert the 3-address CFG to SSA with Cytron's algorithm.
\end{enumerate}

\subsection{From SSA to MPC Source} %: MUX Nodes and Pseudo $\phi$-nodes}

Once the compiler converts the code to SSA,
it transforms $\phi$-nodes that correspond to if-statements into MUX nodes.
From the 3-address CFG conversion step,
$\phi$-nodes corresponding to if-statements will be in a basic block
with the merge condition property.
For example, if $X_3 = \phi(X_1,X_2)$ is in a block with merge condition $C$,
the compiler transforms it into $X_3 = \mathit{MUX}(C, X_1, X_2)$.
Next, the compiler runs the dead code elimination algorithm from Cytron's SSA paper.

Next, the control-flow graph is \emph{linearized} into MPC Source,
which has loops but no if-then-else-statements.
This means that both branches of all if-statements are executed,
and the MUX nodes determine whether to use results from the then-block or from the else-block.
The compiler linearizes the control-flow graph with a variation of breadth-first search.
Blocks with the ``merge condition'' property are only considered
the second time they are visited,
since that will be after both branches of the if-statement are visited. 
(The Python AST naturally gives rise to a translation where each conditional has exactly two targets, 
and each ``merge condition'' block has exactly two incoming edges, a TRUE and a FALSE edge.
Thus, each $\phi$-node has exactly two multiplexer arguments, which dovetails into MUX.
This is in contrast with Cytron's algorithm which operates at the level of the CFG and allows for 
$\phi$-nodes with multiple arguments.)
Each time the compiler visits a block,
it adds the block's statements to the MPC source.
If the block ends in a for-instruction,
the compiler recursively converts the body and code after the loop to MPC source
and adds the for-loop and code after the loop to the main MPC source.
If the block does not end in a for-instruction,
the compiler recursively converts all successor branches to MPC source and
appends these to the main MPC source.

>>>>

% Intro Section
\input{introduction.tex}

% Overview Section
\input{overview.tex}

% Analytical Model Section
\input{model.tex}

% Compiler Framework Section
\input{compiler.tex}

% Vectorization Section
\input{vectorization.tex}

% Results Section
\input{results.tex}

% Related Work Section
\input{related.tex}
